Pool Model Paper

### 2023-12-05 pool_model.tex
more formatting to visually distinguish between ((sub)sub)section

\section{Introduction}

Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}

\section{Motivating Example}

@@

\cmtb{Maybe Example from Seminar by Jens? 2 species,  1 resource, 1 waste production.}

\section{General Mathemetical Formulation}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

@@

@@

Such model can be simplified by explointing that limitation of the resource pool also limits the total bacterial count of the system:
\begin{equation}
    R = 1 - \frac{L+G+D}{N_t}
\end{equation}
and, hence, the full system can be written as 
\begin{equation}
    \begin{cases}
        \dot{L} = -(\mu + \lambda) L\\
        \dot{G} = \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\mu' G\\
        \dot{D} = \mu  L + \mu' G- \beta D 
        \label{eq:3pool_resource} 
    \end{cases}.
\end{equation}
Rewriting the equations above, we get the following non-linear system:
\begin{equation}
    \dot \mbx = \begin{pmatrix}
        -(\mu + \lambda) & 0       & 0 \\
         \lambda         & \alpha  & 0 \\
         \mu & + \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
\end{equation}
Here the first term is the linear part and the second term define the quadratic behaviour of the system.
\begin{figure}[t]
    \begin{center}
    \includegraphics[width=0.6\textwidth]{Figures/pool_model_3pools_resource.pdf}
    \caption{{
        \footnotesize The time dependence of the total bacterial count $N(t) = \sum_i \sum_k \mbx_i^k = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
        All curves were calculated using the model (\ref{eq:3pool_resource}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.
        The purple curve shows significant dip in behaviour of the system where the death of the bacteria is present $\mu = 0.5$.
        Other curves show the case of no bacteria dying $\mu=0$ and represent different length of lag phases:
        short $\lambda=10^{-2}$ (red), intermediate $\lambda=10^{-3}$ (blue) and long one $\lambda=10^{-4}$ (green).
    }}
    \label{fig:3pool_resource_plots}
    \end{center}
\end{figure}
%
%
\subsection{Lag-time Estimation}
Note that one could further simplify this model to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$. 
The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients. 
If yes, the three pool model would be advantageous, if not the two pool model is sufficient with the advantages of having one parameter less.

@@

@@

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=0.67\textwidth]{Figures/pool_model_2pools_resource_competition.pdf}
    \caption{{\footnotesize The competition of the two bacterial species due to the common limiting resource calculated using Eq. (\ref{eq:model_2sp_resource_comp}).
    The red solid curve describes the growth of the bacteria $A$ $N_A = \sum_{k} x_A^k = L_A+G_A$ with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
    The blue solid curve describes the growth of the bacteria $B$ $N_B = \sum_{k} x_B^k = L_B+G_B$ with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.
    The purple dashed line shows the total bacterial count of the system  $N(t)=\sum_i\sum_k\mbx_i^k=N_A+N_B$.
    The gray solid line denotes the dynamic of the resource pool scaled with its capacity $N_t=10^4$.}}
    \label{fig:2pool_resource_2sp}
    \end{center}
\end{figure}
As one can see in Figure~\ref{fig:2pool_resource_2sp} even without direct interaction between species the bacterial growth was inhibited due to the limitation of the resource.
Here species $A$ was able to supress species $B$ due to the higher values of the growth constant $\alpha$.

\cmt{But LV for several bacteria states in literature? Didn't find!}

\subsection{Interspecies Competition}

However, except the inhibition due to resource competition, bacteria developed different others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.

\subsubsection{Toxins Production}
Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $T$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
\begin{eqnarray}
    \dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
    \dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
    \dot{R} &=&-\frac{\alpha}{N_t} R G\\
    \dot{T} &=& \mathcal{F}(R,G).
\end{eqnarray}
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
\begin{eqnarray}
    \mathcal{F}_1(R,G)&=&\kappa G\\
    \mathcal{F}_2(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}
\end{eqnarray}
We would like to test the following model:
\begin{equation}
    \begin{cases}
        \dot{L}_A = - \lambda_A R L_A\\
        \dot{L}_B = - \lambda_B R L_B \\
        \dot{G}_A = \lambda_A R L_A + \alpha_A R G_A - \frac{\mu}{\kappa N_t} T G_A\\
        \dot{G}_B = \lambda_B R L_B + \alpha_B R G_B\\
        \dot{R} = -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B\\
        \dot{T} = \mathcal{F}_{1,2} (R, G_B) \\
    \end{cases}
    \label{eq:model_2sp_toxin}
\end{equation}

@@

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=1.\textwidth]{Figures/pool_model_2pools_toxin.pdf}
    \caption{{\footnotesize The two species system competing for the common nutritient $R$ with production of the toxin $T$ (brown solid line) calculated using Eq. (\ref{eq:model_2sp_toxin}).
    The growing bacteria $B$ produces toxin $T$ (a) constantly $\dot{T} = \mathcal{F}_1 = \kappa G_B$ and (b) only during growth process $\dot{T} = \mathcal{F}_2 = \kappa \alpha_B R G_B$ with 
    production rate $\kappa=0.2$.
    The toxin $T$ causes the cell death of bacteria $A$ with rate constant $\mu = 0.2$.
    The red solid curve describes the growth of the bacteria $A$ $N_A = \sum_{k} x_A^k = L_A+G_A$ with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
    The blue solid curve describes the growth of the bacteria $B$ $N_B = \sum_{k} x_B^k = L_B+G_B$ with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.
    The gray solid line denotes the dynamic of the resource pool scaled with its capacity $N_t=10^4$.}}
    \label{fig:2pool_2sp_toxin}
    \end{center}
\end{figure}

>>>>

\newpage
%
%
%
%
% ================================================================================================
\section{Introduction}

Comprehensive reviews about competition/cooperation bacterial mechanisms: \cite{ghoul_ecology_2016, stubbendieck_bacterial_2016, hibbing_bacterial_2010, west_social_2007}
%
%
%
%
% ================================================================================================
\section{Motivating Example}

@@

\cmtb{Maybe Example from Seminar by Jens? 2 species,  1 resource, 1 waste production.}
%
%
%
%
% ================================================================================================
\section{General Mathemetical Formulation}
Consider the system consisting of $N$ species in $M$ different states.
In the general representation, the system can be described with the Ordinary Differential Equations (\acp{ode}) using a non-linear function $\mbF$:
\begin{equation}
   \dot{\mbx} = \mbF(\mbx, \mbm, \mbp, \mbu),
   \label{eq:model_ODE}
\end{equation}
where  $\mbx \in \mathbb{R}^{N}  \otimes \mathbb{R}^{M}$ is a vector descibing bacterial count.
In matrix form it can be written as
\begin{equation}
    \mbx = \begin{pmatrix}
        x^1_1  & \dots & x^1_N  \\
        \vdots &       & \vdots \\
        x^M_1  & \dots & x^M_N  \\
            \end{pmatrix}
    \label{eq:model_bact}
\end{equation}
with $x_{j}^{i}$ describing the bacterial counts for species $i$ in state $j$.
$\mbu$ is the vector of different inputs of the model, e.g., temperature, pressure etc.
The vector $\mbm \in \mathbb{R}^{K}$ includes all variables corresponding to the microenvironment of the system which can be concentartion of the inhibitors or activators:
\begin{equation}
    \dot{\mbm} = \mbH (\mbx, \mbm, \mbp, \mbu).
    \label{eq:model_microenv}
\end{equation}
Here function $\mbH$ is in general case also non-linear.

@@

% ================================================================================================

@@

% ------------------------------------------------------------------------------------------------

@@

Such model can be simplified by explointing that limitation of the resource pool also limits the total bacterial count of the system:
    R = 1 - \frac{L+G+D}{N_t}
and, hence, the full system can be written as 
        \dot{G} = \lambda L + \alpha G\left(1-\frac{L+G+D}{N_t}\right)-\mu' G\\
        \label{eq:3pool_resource} 
Rewriting the equations above, we get the following non-linear system:
         \lambda         & \alpha  & 0 \\
         \mu & + \mu'& - \beta 
    \end{pmatrix} 
    \mbx + \frac{\alpha}{N_t} x^2 \sum_i^3 x^i \begin{pmatrix} 0 \\ 1 \\ 0  \end{pmatrix}.
Here the first term is the linear part and the second term define the quadratic behaviour of the system.
    \includegraphics[width=0.6\textwidth]{Figures/pool_model_3pools_resource.pdf}
    \caption{{
        \footnotesize The time dependence of the total bacterial count $N(t) = \sum_i \sum_k \mbx_i^k = L(t) + G(t) + D(t)$ is obtained as a sum of bacteria in all states.
        All curves were calculated using the model (\ref{eq:3pool_resource}) with parameters $\alpha=2.5$, $\mu'=0.5$ and $\beta=5$.
        The purple curve shows significant dip in behaviour of the system where the death of the bacteria is present $\mu = 0.5$.
        Other curves show the case of no bacteria dying $\mu=0$ and represent different length of lag phases:
        short $\lambda=10^{-2}$ (red), intermediate $\lambda=10^{-3}$ (blue) and long one $\lambda=10^{-4}$ (green).
    \label{fig:3pool_resource_plots}
% ------------------------------------------------------------------------------------------------
\subsection{Lag-time Estimation}
Note that one could further simplify this model to a two pool or state model by omitting the pool $D$ and removing cells with rate $\beta$ directly from pool $L$. 
The relevant question here is whether cells in the pool $D$ still contribute to the limitation of the resources by, e.g., occupying space or consuming important nutrients. 
If yes, the three pool model would be advantageous, if not the two pool model is sufficient with the advantages of having one parameter less.

@@

% ................................................................................................

@@

\begin{figure}[H]
    \begin{center}
    \includegraphics[width=0.67\textwidth]{Figures/pool_model_2pools_resource_competition.pdf}
    \caption{{\footnotesize The competition of the two bacterial species due to the common limiting resource calculated using Eq. (\ref{eq:model_2sp_resource_comp}).
    The red solid curve describes the growth of the bacteria $A$ $N_A = \sum_{k} x_A^k = L_A+G_A$ with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
    The blue solid curve describes the growth of the bacteria $B$ $N_B = \sum_{k} x_B^k = L_B+G_B$ with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.
    The purple dashed line shows the total bacterial count of the system  $N(t)=\sum_i\sum_k\mbx_i^k=N_A+N_B$.
    The gray solid line denotes the dynamic of the resource pool scaled with its capacity $N_t=10^4$.}}
    \label{fig:2pool_resource_2sp}
    \end{center}
\end{figure}
As one can see in Figure~\ref{fig:2pool_resource_2sp} even without direct interaction between species the bacterial growth was inhibited due to the limitation of the resource.
Here species $A$ was able to supress species $B$ due to the higher values of the growth constant $\alpha$.
%
\cmt{But LV for several bacteria states in literature? Didn't find!}
%
%
%
% ------------------------------------------------------------------------------------------------
\subsection{Interspecies Competition}
However, except the inhibition due to resource competition, bacteria developed different others active competition mechanisms.
One of the examples of such mechanisms is a production of the antimicrobial compounds, inhibitors or toxins~\cite{wloch-salamon_effect_2008, chao_structured_1981}.
%
%
% ................................................................................................
\subsubsection{Toxins Production}
Now extend our model where bacteria not only consume nutritions from resource pool $R$ but one of the species also actively produce the inhibiting compounds $T$. 
It is the accumulation of the bacteria waste which is responsible for the spoilage of food products. 
In order to take this into account we add an equation for the waste production:
\begin{eqnarray}
    \dot{L} &=& - \lambda_f R L + \mathcal{B}(R)G\\
    \dot{G} &=& \lambda_f R L +\alpha R G-\mathcal{B}(R)G\\
    \dot{R} &=&-\frac{\alpha}{N_t} R G\\
    \dot{T} &=& \mathcal{F}(R,G).
\end{eqnarray}
The function $\mathcal{F}$ denotes the model for the antibiotic production rate. 
We shall discuss two possible models:
\begin{eqnarray}
    \mathcal{F}_1(R,G)&=&\kappa G\\
    \mathcal{F}_2(R,G)&=&\kappa\alpha R G=-\kappa N_t\dot{R}
\end{eqnarray}
We would like to test the following model:
\begin{equation}
    \begin{cases}
        \dot{L}_A = - \lambda_A R L_A\\
        \dot{L}_B = - \lambda_B R L_B \\
        \dot{G}_A = \lambda_A R L_A + \alpha_A R G_A - \frac{\mu}{\kappa N_t} T G_A\\
        \dot{G}_B = \lambda_B R L_B + \alpha_B R G_B\\
        \dot{R} = -\frac{\alpha_A}{N_t} R G_A-\frac{\alpha_B}{N_t} R G_B\\
        \dot{T} = \mathcal{F}_{1,2} (R, G_B) \\
    \end{cases}
    \label{eq:model_2sp_toxin}
\end{equation}

@@

    \includegraphics[width=1.\textwidth]{Figures/pool_model_2pools_toxin.pdf}
    \caption{{\footnotesize The two species system competing for the common nutritient $R$ with production of the toxin $T$ (brown solid line) calculated using Eq. (\ref{eq:model_2sp_toxin}).
    The growing bacteria $B$ produces toxin $T$ (a) constantly $\dot{T} = \mathcal{F}_1 = \kappa G_B$ and (b) only during growth process $\dot{T} = \mathcal{F}_2 = \kappa \alpha_B R G_B$ with 
    production rate $\kappa=0.2$.
    The toxin $T$ causes the cell death of bacteria $A$ with rate constant $\mu = 0.2$.
    The red solid curve describes the growth of the bacteria $A$ $N_A = \sum_{k} x_A^k = L_A+G_A$ with model parameters $\lambda_A=10^{-2}$, $\alpha_A=3.0$.
    The blue solid curve describes the growth of the bacteria $B$ $N_B = \sum_{k} x_B^k = L_B+G_B$ with model parameters $\lambda_A=5\cdot 10^{-2}$, $\alpha_A=2.5$.
    The gray solid line denotes the dynamic of the resource pool scaled with its capacity $N_t=10^4$.}}
    \label{fig:2pool_2sp_toxin}
% ................................................................................................
